---
format: 
  revealjs:
    css: ../../styles.css
    slide-number: true
    show-slide-number: all
    progress: true
    history: true
    hash-type: number
    theme: default
    code-block-background: true
    highlight-style: github
    code-link: false
    code-copy: true
    controls: true
    pagetitle: "Inter R4SS Day 3B"
    author-meta: "Jeffrey Girard"
    date-meta: "2023-06-07"
---

::: {.my-title}
# [Intermediate R]{.blue2} <br />for Social Scientists

::: {.my-grey}
[Workshop Day 3B | 2023-06-07]{}<br />
[Jeffrey M. Girard | Pitt Methods]{}
:::

![](../../img/proud2_2780E3.svg){.absolute bottom=0 right=0 width=400}
:::

# Generalized <br />Linear Modeling

## The Linear Model (LM)

$$
\begin{align}
\mu_i &= b_0 + b_1 x_{1i} + \cdots + b_m x_{mi}\\
y_i &\sim \text{Gaussian}(\mu_i, \sigma)
\end{align}
$$

::: {.fragment .f80}

- The first (structural) term is [linear]{.b .blue} (i.e., the sum of products)
    -   It assumes that $\mu_i$ can vary indefinitely, i.e., $y \in (-\infty,\infty)$
    
:::

::: {.fragment .f80}
- The second (random) term describes the [shape of the residuals]{.b .blue}
    -   It assumes they are [normally (Gaussian) distributed]{.b .green} around $\mu_i$
    -   It assumes they have [constant variance]{.b .green} (i.e., there is only one $\sigma$)
    
:::

## The Generalized LM (GLM)

::: {.f80}
- What happens when these assumptions are violated?
    -   Our parameters and inferences will be biased
    -   Our model may make impossible predictions

:::

::: {.fragment}

::: {.f80}
- GLM addresses this issue by making two changes
    -   Adds a [family]{.b .blue}, i.e., specifies a different shape of the residuals
    -   Adds a [link function]{.b .green}, i.e., transforms $y$ to vary indefinitely

:::

$$
\begin{align}
\color{ForestGreen}{\text{link}}(\mu_i) &= b_0 + b_1 x_{1i} + \cdots + b_m x_{mi} \\
y_i &\sim \color{blue}{\text{family}}(\mu_i, \ldots)
\end{align}
$$
:::


## Basic GLM Families

- [Normal:]{.b .green} $y \in \mathbb{R}$, use `gaussian()` with identity link
- [Nonnegative:]{.b .green} $y \in \mathbb{R}^+$, use `Gamma()` with inverse link
- [Binary:]{.b .blue} $y \in \{0,1\}$, use `binomial()` with logit/probit link
- [Counts:]{.b .blue} $y \in \{0, 1, 2, ...\}$, use `poisson()` with log link

::: {.fragment .mt1}
-   Many advanced/specialized GLM families are also available!
:::

## Binary Regression

-   Binary regression is used when $y$ is binary, i.e., $0$ or $1$
    -   *e.g., Is the patient healthy (0) or sick (1)?*
    -   *e.g., Did the student fail (0) or pass (1)?*
    -   *e.g., Is the attribute absent (0) or present (1)?*
    -   *e.g., Was the coin flip heads (0) or tails (1)?*

## The Binomial Family

$$
k \sim \text{Binomial}(n,p)
$$

- $n\in\{0,1,2,...\}$ --- the number of trials
- $p\in[0,1]$ --- the probability of success in each trial
- $k\in\{0,1,...,n\}$ --- the number of successes

::: {.fragment}
-   This family applies to [binary]{.b .blue} cases when $n=1$
    -   *e.g., If I flip one coin, how many landed heads?*
    
:::

## Logistic Regression

$$
\begin{align}
\text{logit}(p_i) &= b_0 + b_1 x_{1i} + \cdots + b_m x_{mi} \\
y_i &\sim \text{Binomial}(1, p_i)
\end{align}
$$

-   We can't predict $p$ directly because it is bounded, $p\in[0,1]$
-   The [logit link]{.b .blue} makes $p$ unbounded, $\text{logit}(p)\in(-\infty,\infty)$

$$\text{logit}(p)=\log(\text{Odds}(p))=\log\left(\frac{p}{1-p}\right)$$

## The Logit Link

```{r}
#| echo: false
#| message: false

library(tidyverse)
tibble(
  p = seq(0.001, 0.999, by = 0.001),
  `logit(p)` = log(p / (1 - p))
) |> 
  ggplot(aes(y = p, x = `logit(p)`)) +
  geom_vline(
    color = "white", 
    linewidth = 4, 
    xintercept = 0
  ) +
  geom_line(color = "royalblue", linewidth = 2) +
  scale_x_continuous(breaks = seq(-6, 6, 2)) +
  theme_gray(base_size = 24) +
  theme(panel.grid.minor = element_blank())

```


## Interpretation

- Our slopes ($b_m$) will be in [logit (log-odd) units]{.b .green}
    -   $b_m>0$ indicates increased prob., $b_m<0$ decreased

::: {.fragment}
- We can exponentiate them to convert to [Odds Ratios]{.b .blue}
    -   If $b_1=0.3$, then $OR=e^{0.3}=1.35$ times the odds of success (or a 35% increase) when increasing $x_1$ by 1
    -   If $b_1=-0.3$, then $OR=e^{-0.3}=0.74$ times the odds of success (or a 26% decrease) when increasing $x_1$ by 1

:::

## Binary Regression Live Coding

```{r}
#| echo: true
#| eval: false
#| error: true
#| code-line-numbers: false

## Load packages (after installing, if needed)
library(tidyverse)
library(easystats)

## Read in some example binary data and mutate the outcome to be 0 vs. 1
titanic <- 
  read_csv("titanic.csv") |> 
  mutate(survived = if_else(survived == "yes", true = 1, false = 0)) |> 
  print()

## Visualize the relationship between fare and survival
ggplot(titanic, aes(x = fare, y = survived)) +
  geom_point() +
  geom_smooth()

## Fit a basic regression using LM
fit <- lm(
  formula = survived ~ fare, 
  data = titanic
)
model_parameters(fit)
model_parameters(fit) |> print_md() #looks better in quarto
model_performance(fit)
plot(estimate_expectation(fit))
check_model(fit)

## Fit a logistic regression using GLM
fit2 <- glm(
  formula = survived ~ fare, 
  family = binomial(link = "logit"), 
  data = titanic
)
model_parameters(fit2)
model_parameters(fit2, exponentiate = TRUE)
model_performance(fit2)
plot(estimate_link(fit2))
plot(estimate_expectation(fit2))
check_model(fit2)

## Estimate the model's expectation for specific values of the predictor(s)
estimate_expectation(fit2, tibble(fare = c(10, 50, 100)))

## Note: To fit a probit binary regression, just use binomial(link = "probit")
```


## Count Regression

- Count regression is used when $y$ is a [whole number]{.b .green}

::: {.fragment}
- Counts can be [bounded]{.b .green} (i.e., have a maximum number)
    - *e.g., how many answers on the test were correct?*
    - For these, we can re-use the **Binomial** family $(n>0)$

:::

::: {.fragment}
- Counts can also be [unbounded]{.b .blue} (i.e., no maximum)
    - *e.g., how many hospitalizations in the last 10 years?*
    - For these, we can use **Poisson** or **Negative Binomial**
    
:::


## The Poisson Family

$$
k \sim \text{Poisson}(\lambda)
$$

- $\lambda\in(0,\infty)$ --- occurrence rate
- $k\in\mathbb{N}_0$ --- number of occurrences

```{r, fig.width=10, fig.height=3}
#| echo: false

tibble(`λ` = c(1, 2, 4, 8)) |>
  mutate(k = map(`λ`, \(x) rpois(1000, x))) |> 
  unnest(k) |> 
  ggplot(aes(x = k)) + 
  geom_bar() +
  facet_wrap(~`λ`, labeller = label_both, scales = "free", nrow = 1) +
  # scale_x_continuous(breaks = seq(0, 20, 4)) +
  theme_grey(base_size = 22) +
  theme(panel.grid.minor = element_blank())

```


## Poisson Regression
$$
\begin{align}
\log(\lambda_i) &= b_0 + b_1 x_{1i} + \cdots + b_m x_{mi} \\
y_i &\sim \text{Poisson}(\lambda_i)
\end{align}
$$

-   We can't predict $\lambda$ directly; it is bounded $\lambda\in(0,\infty)$
-   The [log link]{.b .blue} makes $\lambda$ unbounded, $\log(\lambda)\in(-\infty,\infty)$

## Interpretation

- Our slopes ($b_m$) will be in [log units]{.b .green}
    -   $b_m>0$ indicates increased rate, $b_m<0$  decreased

::: {.fragment}
- We can exponentiate to convert to [Incidence Rate Ratios]{.b .blue}
    -   If $b_1=0.3$, then $IRR=e^{0.3}=1.35$ times the incidence rate (a 35% increase) when increasing $x_1$ by 1
    -   If $b_1=-0.3$, then $IRR=e^{-0.3}=0.74$ times the incidence rate (a 26% decrease) when increasing $x_1$ by 1

:::

## Count Regression Live Coding

```{r}
#| echo: true
#| eval: false
#| error: true
#| code-line-numbers: false

# Read in some example count data
disc <- read_csv("discoveries.csv")
disc

# Visualize the relationship between year and discovery count
ggplot(disc, aes(x = year, y = count)) + 
  geom_point() +
  geom_smooth()

# Fit a linear regression using LM
fit <- lm(
  formula = count ~ poly(year, degree = 2), 
  data = disc
)
model_parameters(fit)
model_performance(fit)
plot(estimate_relation(fit))
check_model(fit)

# Fit a poisson regression using GLM
fit2 <- glm(
  formula = count ~ poly(year, degree = 2),
  family = poisson(link = "log"), 
  data = disc
)
model_parameters(fit2)
model_parameters(fit2, exponentiate = TRUE)
model_performance(fit2)
plot(estimate_relation(fit2))

# Advanced: Overdispersion and quasi-poisson
check_overdispersion(fit2)
fit3 <- glm(
  count ~ poly(year, degree = 2), 
  family = quasipoisson(link = "log"), 
  data = disc
)
model_parameters(fit3)
model_parameters(fit3, exponentiate = TRUE)
plot(estimate_relation(fit3))

# Compare parameters between poisson and quasi-poisson
compare_parameters(fit2, fit3, select = "ci_p2")
```

# Multilevel (Mixed Effects) Modeling

## Conceptual Overview {.smaller}

- LM and GLM both assume [independent residuals]{.b .blue}
    - This is violated by clustered and longitudinal data
    - The estimated SEs will be **biased** (usually too small)

::: {.fragment}
- [Clustered data]{.b .green} has a higher level of organization
    - *e.g., students from same classroom are more similar*
    - *e.g., participants from same country are more similar*
    
:::

::: {.fragment}
- [Longitudinal data]{.b .green} is repeated sampled from same source
    - *e.g., each participant responds to multiple trials*
    - *e.g., the global temperature is measured every week*

:::

## Cluster-Robust Standard Errors {.smaller}

- The basic OLS standard errors will be biased with clustered data
- CR-SEs use a statistical procedure to try to correct this bias
    -   This is convenient when clustering is just a "nuisance"

::: {.fragment}
1. Estimate the regression coefficients using OLS (these are unchanged)
1. Estimate the CR-SEs using a more complex formula, e.g., based on clusters $J$
1. Interpret the coefficients and CR-SEs as you would in a basic LM or GLM

:::

::: {.fragment .tc}
$$
\begin{align}
Var^{OLS}(\hat{b})&=\sigma^2(\mathbf{X}^\intercal\mathbf{X})^{-1} \\
Var^{CR}(\hat{b})&=(\mathbf{X}^\intercal\mathbf{X})^{-1}\sum_{j=1}^J(\mathbf{X}_j^\intercal e_je_j^\intercal \mathbf{X}_j)(\mathbf{X}^\intercal \mathbf{X})^{-1}
\end{align}
$$

*In the latter, we sum residuals by cluster instead of observation*
:::

## Cluster-Robust Standard Errors

```{r}
#| echo: true
#| eval: false
#| error: true
#| code-line-numbers: false

library(tidyverse)
library(easystats)
library(sandwich)
mlmath <- read_csv("mlmath.csv")
mlmath

mlmath |> count(school)
mlmath |> count(public)

# Fit LM ignoring clustering within schools

fit <- lm(
  formula = math ~ 1 + homework,
  data = mlmath
)
model_parameters(fit)
model_performance(fit)
estimate_relation(fit) |> plot()

# Estimate cluster-robust standard errors

model_parameters(
  fit, 
  vcov = "vcovCL", 
  vcov_args = list(type = "HC1", cluster = mlmath$school)
)
estimate_relation(
  fit,
  vcov = "vcovCL", 
  vcov_args = list(type = "HC1", cluster = mlmath$school)
) |> 
  plot()
```

## Simpson's Paradox {.smaller}

However, reaching the right conclusions may require knowing the clusters...

```{r}
#| echo: false

set.seed(2021)
d1 <- 
  tibble(
    hospital = rep(1:3, each = 40),
    symptoms = c(rnorm(40, 3, 0.75), rnorm(40, 5, 0.75), rnorm(40, 7, 0.75)),
    prognosis = 3 + 1.5*hospital - 0.5*symptoms + rnorm(120, 0, 0.25)
  ) |> 
  mutate(
    hospital = factor(hospital, labels = c("Children's", "Mercy", "University"))
  )

ggplot(d1, aes(x = symptoms, y = prognosis)) + 
  geom_point(size = 3) +
  scale_y_continuous(limits = c(1.5, 5.5)) +
  theme_grey(base_size = 24)
```

## Simpson's Paradox {.smaller}

Here it looks like having **worse** symptoms predicts a ***better*** prognosis...

```{r}
#| echo: false

ggplot(d1, aes(x = symptoms, y = prognosis)) + 
  geom_point(size = 3) + 
  geom_smooth(formula = y~x, method = "lm", linewidth = 1) + 
  scale_y_continuous(limits = c(1.5, 5.5)) +
  theme_grey(base_size = 24)
```

## Simpson's Paradox {.smaller}

But this data actually comes from three different hospitals!

```{r, echo=FALSE}
ggplot(
  d1, 
  aes(x = symptoms, y = prognosis, color = hospital, shape = hospital)
) +
  geom_point(size = 3) + 
  scale_y_continuous(limits = c(1.5, 5.5)) +
  theme_grey(base_size = 24)
```

## Simpson's Paradox {.smaller}

Within each hospital, the relationship is the opposite!

```{r, echo=FALSE}
ggplot(
  d1, 
  aes(x = symptoms, y = prognosis, color = hospital, shape = hospital)
) + 
  geom_point(size = 3) + 
  geom_smooth(formula = y~x, method = "lm", linewidth = 1) + 
  scale_y_continuous(limits = c(1.5, 5.5)) +
  theme_grey(base_size = 24)
```

## Levels of Clustering {.smaller}

- We can think and talk about clustered data as having [multiple levels]{.b .blue} (hence MLM)

::: {.fragment}
- We number levels from $1$ to $M$ where lower numbers represent more specific levels
    - *e.g., Data is collected from students (L1) nested within classrooms (L2)*
    - *e.g., Data is collected from students (L1) in classrooms (L2) in schools (L3)*
    - *e.g., Data is collected from timepoints (L1) nested within participants (L2)*
    - *e.g., Data is collected from timepoints (L1) in participants (L2) in countries (L3)*

:::

::: {.fragment}
- What is and isn't considered a level?
    - At each level, we are sampling from a broader population
    - Thus, levels need to be *random* variables (and are usually discrete)
    - *If we repeated the study, would the variable's values be the same?*
    
:::

## Variables at Different Levels

- We can have variables that vary at (or describe) each level
    - L1: each **student**'s `age`, `sex`, and `test_score`
    - L2: each **classroom**'s `subject` and `n_students`
    - L3: each **school**'s `budget` and `charter_status`

::: {.fragment}
- In standard MLM, the outcome/DV has to be on L1
    - But we can still use higher level predictors
    - We can even have cross-level interactions

:::

## Mixed Effects (Multilevel) Modeling {.smaller}

When we have multiple clusters of data, we have several options for fitting linear models

1. **Complete Pooling:** Fit a single line to all the data, ignoring clustering

1. **No Pooling:** Fit a separate line for each cluster, ignoring any similarity

1. **Partial Pooling:** Fit a *distribution* of similar lines to the data, one for each cluster

::: {.fragment .mt2}
Multilevel modeling uses the partial pooling approach, which yields two types of effects

- [Fixed Effects]{.b .green} either describe an effect assumed to be the same for all clusters, <br />or describe the center of an effect's distribution across clusters

- [Random Effects]{.b .blue} describe how each cluster deviates from the fixed effect and therefore create the distribution across clusters (can be estimated or not)

:::

## Example Models {.smaller}

0. **Null Model**
    - Each cluster (school) has its own intercept (average math score)
    - There are no predictors/slopes added (fixed or random)

::: {.fragment}
1. **Random Intercepts (RI) Model**
    - Each cluster (school) has its own intercept (average math score)
    - The slope (homework effect) is fixed and applies to all clusters (schools)
    
:::

## Example Models {.smaller}

2. **Random Intercepts and Slopes (RIS) Model**
    - Each cluster (school) has its own intercept (average math score)
    - Each cluster (school) has its own slope (homework effect)

::: {.fragment}
3. **Cross-Level Interaction Model**
    - An L1 (student) outcome will be regressed on an L2 (school) predictor
    - An L1 (student) predictor will interact with an L2 (school) predictor
    
:::

## Null and RI Models Live Coding

```{r}
#| echo: true
#| eval: false
#| error: true
#| code-line-numbers: false

library(lme4)
library(merDeriv)
library(ggeffects)

# ==============================================================================
# Fit and examine null model

fit_null <- lmer(
  formula = math ~ 1 + (1 | school),
  data = mlmath
)
model_parameters(fit_null) |> print_md()
model_performance(fit_null) |> print_md()

# ==============================================================================
# Fit and examine random intercepts model

fit_ri <- lmer(
  formula = math ~ 1 + homework + (1 | school), 
  data = mlmath
)
model_parameters(fit_ri) |> print_md()
model_performance(fit_ri) |> print_md()

plot(estimate_relation(fit_ri))

## random effects (deviations from the fixed effect)
estimate_grouplevel(fit_ri) |> print_md() 
estimate_grouplevel(fit_ri) |> plot()

## fixed effects + random effects (deviation)
estimate_grouplevel(fit_ri, type = "total") |> print_md() 

## plot the line for each cluster (school)
ggpredict(fit_ri, c("homework", "school"), type = "random") |> 
  plot(colors = 1:10, ci = FALSE)
```

## RIS and CLI Models Live Coding

```{r}
#| echo: true
#| eval: false
#| error: true
#| code-line-numbers: false

# ==============================================================================
# Fit and examine random intercepts and slopes model

fit_ris <- lmer(
  formula = math ~ 1 + homework + (1 + homework | school),
  data = mlmath
)
model_parameters(fit_ris) |> print_md()
model_performance(fit_ris) |> print_md()

plot(estimate_relation(fit_ris))

## random effects (deviations from the fixed effect)
estimate_grouplevel(fit_ris) |> print_md() 
estimate_grouplevel(fit_ris) |> plot()

## fixed effects + random effects (deviation)
estimate_grouplevel(fit_ris, type = "total") |> print_md() 

## plot the line for each cluster (school)
ggpredict(fit_ris, c("homework", "school"), type = "random") |> 
  plot(colors = 1:10, ci = FALSE)

## turn refit off (use REML) to compare models with the same fixed effects
anova(fit_ri, fit_ris, refit = FALSE)

# ==============================================================================
# Fit and examine the cross-level interaction model

fit_cli <- lmer(
  formula = math ~ 1 + homework * public + (1 + homework | school),
  data = mlmath
)
model_parameters(fit_cli) |> print_md()
model_performance(fit_cli) |> print_md()

plot(estimate_relation(fit_cli))

## estimate simple slopes
estimate_slopes(fit_cli, trend = "homework", at = "public") |> print_md()

## Turn refit on (use ML) to compare models with different fixed effects
anova(fit_ris, fit_cli, refit = TRUE)
```

## Centering

```{r}
#| echo: true
#| eval: false
#| error: true
#| code-line-numbers: false

# Grand mean centering (good for level-2 variables)

mlmath2 <- 
  mlmath |> 
  mutate(ratio_c = center(ratio)) |> 
  print()

# Centering within clusters (good for level-1 variables)

mlmath3 <- 
  mlmath2 |> 
  mutate(
    homework_m = mean(homework), # cluster mean (between)
    homework_d = homework - homework_m, # deviation from cluster mean (within)
    .by = school
  ) |> 
  print()

## Incorporating these variables into the model

fit <- lmer(
  math ~ 1 + homework_m + homework_d + ratio_c + (1 + homework_d | school),
  data = mlmath3
)
model_parameters(fit) |> print_md()
model_performance(fit) |> print_md()
```


## Longitudinal Considerations

```{r}
#| echo: true
#| eval: false
#| error: true
#| code-line-numbers: false

# ==============================================================================
# Load example data

soccer <- 
  read_csv("soccer.csv") |> 
  mutate(across(c(Code, Played), factor)) |> 
  print()

# ==============================================================================
# Fit null model and examine ICC

fit0 <- lmer(
  Testosterone ~ 1 + (1 | Code),
  data = soccer
)
model_performance(fit0) |> print_md()

# ==============================================================================
# Fit random intercepts longitudinal model

fit1 <- lmer(
  Testosterone ~ 1 + time0 + (1 | Code),
  data = soccer
)
model_parameters(fit1) |> print_md()
model_performance(fit1) |> print_md()

estimate_relation(fit1) |> plot()

ggpredict(fit1, c("time0", "Code"), type = "random") |> 
  plot(colors = 1:25, ci = FALSE)

# ==============================================================================
# Fit random intercepts and slopes model

fit2 <- lmer(
  Testosterone ~ 1 + time0 + (1 + time0 | Code),
  data = soccer
)
model_parameters(fit2) |> print_md()
model_performance(fit2) |> print_md()

ggpredict(fit2, c("time0", "Code"), type = "random") |> 
  plot(colors = 1:25, ci = FALSE)

## Turn refit off (use REML) when comparing models with same fixed effects
anova(fit1, fit2, refit = FALSE)

# ==============================================================================
# Fit cross-level interaction model

fit3 <- lmer(
  Testosterone ~ 1 + time0 * Played + (1 + time0 | Code),
  data = soccer
)
model_parameters(fit3) |> print_md()
model_performance(fit3) |> print_md()

estimate_relation(fit3) |> plot()

ggpredict(fit3, c("time0", "Code", "Played"), type = "random") |> 
  plot(colors = 1:25, ci = FALSE)

## Turn refit on (use ML) when comparing models with different fixed effects
anova(fit2, fit3, refit = TRUE)
```

