---
format: 
  revealjs:
    css: ../../styles.css
    slide-number: true
    show-slide-number: all
    progress: true
    history: true
    hash-type: number
    theme: default
    code-block-background: true
    highlight-style: github
    code-link: false
    code-copy: true
    controls: true
    pagetitle: "Inter R4SS Day 2A"
    author-meta: "Jeffrey Girard"
    date-meta: "2023-06-06"
---

::: {.my-title}
# [Intermediate R]{.blue2} <br />for Social Scientists

::: {.my-grey}
[Workshop Day 2A | 2023-06-06]{}<br />
[Jeffrey M. Girard | Pitt Methods]{}
:::

![](../../img/proud2_2780E3.svg){.absolute bottom=0 right=0 width=400}
:::

# Summarize

## Summarize {.smaller}

::: {.columns .pv4}

::: {.column width="60%"}
-   Although we store data about many observations...
-   ...we often want to [summarize across]{.b .blue} observations
    -   This is like [folding]{.b .green} the tibble down to [one row]{.b .green}

::: {.fragment .mt1}
-   We've seen functions that summarize vectors
    -   `length()`, `sum()`, `min()`, `max()`
    -   `mean()`, `median()`, `sd()`, `var()`
:::

::: {.fragment .mt1}
-   `summarize()` lets us use them on tibbles
    -   It works very similarly to `mutate()`
    -   It always creates a tibble as output
:::
:::

::: {.column .tc .pv4 width="40%"}
{{< li uhmqrsgl trigger=loop delay=3000 colors=secondary:#2a76dd class=rc >}}
:::

:::

## Summarize Live Coding

```{r}
#| echo: true
#| eval: false
#| error: true
#| code-line-numbers: false

# SETUP: We will need tidyverse and an example dataset

library(tidyverse)

sales <- 
  tibble(
    customer = c(1, 2, 3, 1, 3),
    store = c("A", "A", "A", "B", "B"),
    items = c(25, 20, 16, 10, 5),
    spent = c(685, 590, 392, 185, 123)
  ) |> 
  print()

# ==============================================================================

# USECASE: Summarize the typical sales

my_summary <- 
  sales |> 
  summarize(
    avg_items = mean(items),
    avg_spent = mean(spent)
  ) |> 
  print()

# ==============================================================================

# PITFALL: Don't use summary() instead of summarize()

my_summary <- 
  sales |> 
  summary(
    avg_items = mean(items),
    avg_spent = mean(spent)
  ) |> 
  print() # not a tibble

# ==============================================================================

# USECASE: Use more than one summary function

my_summary <- 
  sales |> 
  summarize(
    total_items = sum(items),
    total_spent = sum(spent),
    avg_items = mean(items),
    avg_spent = mean(spent)
  ) |> 
  print()

# ==============================================================================

# USECASE: Use counting functions

my_counts <- 
  sales |> 
  summarize(
    n_sales = n(),
    n_customers = n_distinct(customer),
    n_stores = n_distinct(store)
  ) |> 
  print()
```

# Grouped Summarize

## Grouped Summarize {.smaller}

::: {.columns .pv4}

::: {.column width="60%"}
-   We can also summarize a tibble [by group]{.b .blue}
    -   This is like [folding]{.b .green} the tibble multiple times
    -   Specifically, we fold down to [one row per group]{.b .green}

::: {.fragment .mt1}
-   We just add the `.by` argument to `summarize()`
    -   We can group by one or many variables
    -   With many, we group by their combinations
:::
:::

::: {.column .tc .pv4 width="40%"}
{{< li pmwtbpap trigger=loop delay=3000 colors=secondary:#2a76dd class=rc >}}
:::

:::

## Grouped Summarize Live Coding

```{r}
#| echo: true
#| eval: false
#| error: true
#| code-line-numbers: false

# SETUP: We will need tidyverse and an example dataset

library(tidyverse)

sales <- 
  tibble(
    customer = c(1, 2, 3, 1, 3),
    store = c("A", "A", "A", "B", "B"),
    items = c(25, 20, 16, 10, 5),
    spent = c(685, 590, 392, 185, 123)
  ) |> 
  print()

# ==============================================================================

# USECASE: Summarize per store

sales |> 
  summarize(
    customers = n_distinct(customer),
    items_sold = sum(items),
    total_sales = sum(spent),
    avg_items = mean(items),
    avg_spent = mean(spent),
    .by = store
  )

# ==============================================================================

# USECASE: Summarize per customer

sales |> 
  summarize(
    n_visits = n(),
    items_bought = sum(items),
    total_spent = sum(spent),
    stores_visited = n_distinct(store),
    .by = customer
  )

# ==============================================================================

# USECASE: Summarize per customer-store combination

sales |> 
  summarize(
    n_visits = n(),
    items_bought = sum(items),
    total_spent = sum(spent),
    .by = c(customer, store)
  )

# ==============================================================================

# SETUP: Let's get a larger, more realistic dataset

# Extra pane > Packages tab > Install > nycflights13

library("nycflights13")

flights

# ==============================================================================

# USECASE: Find the carrier with the lowest average delays

flights |> 
  summarize(
    m_delay = mean(dep_delay, na.rm = TRUE),
    .by = carrier
  ) |> 
  arrange(m_delay)

# ==============================================================================

# LESSON: We can also group by multiple variables

# USECASE: Let's find the day of the year with the most flights

flights |> 
  summarize(
    n_flights = n(),
    .by = c(month, day)
  ) |> 
  arrange(desc(n_flights))

# ==============================================================================

# Note how this is different from summarizing by day only (365 vs. 31)

flights |> 
  summarize(
    n_flights = n(),
    .by = day
  ) |> 
  arrange(desc(n_flights))
```

# Grouped Mutate

## Grouped Mutate {.smaller}

::: {.columns .pv4}

::: {.column width="60%"}
-   We can also [mutate]{.b .blue} a tibble by group
    -   Adds the summary values in a new column
    -   They will be recycled for groups with many rows
    
::: {.fragment .mt1}
-   We now have four useful tools for different jobs
    -   Ungrouped summarize: fold to one row overall
    -   Grouped summarize: fold to one row per group
    -   Ungrouped mutate: add summaries overall
    -   Grouped mutate: add summaries per group
:::
:::

::: {.column .tc .pv4 width="40%"}
{{< li qygayvwp trigger=loop delay=3000 colors=secondary:#2a76dd class=rc >}}
:::

:::

## Grouped Mutate Live Coding

```{r}
#| echo: true
#| eval: false
#| error: true
#| code-line-numbers: false

sales

# Calculates the total sum (across both stores)
sales |> 
  summarize(
    items_sum_total = sum(items)
  )

# Calculates the store sums
sales |> 
  summarize(
    items_sum_store = sum(items),
    .by = store
  )

# Appends and recycles the total sum (across both stores)
sales |> 
  mutate(
    items_sum_total = sum(items),
    items_pct_total = items / items_sum_total
  )

# Appends and recycles the store sums
sales |> 
  mutate(
    items_sum_store = sum(items),
    items_pct_store = items / items_sum_store,
    .by = store
  )

# What if you want both? You can!

sales |> 
  mutate(
    items_sum_total = sum(items),
    items_pct_total = items / items_sum_total
  ) |> 
  mutate(
    items_sum_store = sum(items),
    items_pct_store = items / items_sum_store,
    .by = store
  )

flights 

# .by can also be used with mutate()
flights |> 
  select(year:dep_time, carrier, dep_delay) |> 
  mutate(
    mean_dep_delay = mean(dep_delay, na.rm = TRUE),
    .by = carrier
  )

# Compare this to summarize version
flights |> 
  select(year:dep_time, carrier, dep_delay) |> 
  summarize(
    mean_dep_delay = mean(dep_delay, na.rm = TRUE),
    .by = carrier
  )

# Example longitudinal data
dat <- 
  tibble(
    person = rep(101:103, each = 4),
    sex = rep(c("M", "F", "M"), each = 4),
    age = rep(c(13, 10, 12), each = 4), 
    visit = rep(1:4, times = 3),
    score = round(rnorm(12), digits = 1)
  ) |> 
  print()

# Another grouped mutate vs grouped summarize example
dat |> 
  mutate(
    n_visits = n(), 
    .by = person
  )

dat |> 
  summarize(
    n_visits = n(),
    .by = person
  )

# Add row numbers with row_number()
dat2 <- 
  dat |> 
  select(-visit) |> 
  print()

dat2 |> 
  mutate(
    visit_total = row_number()
  )

dat2 |> 
  mutate(
    visit_person = row_number(),
    .by = person
  )
```

# Tidy Control Flow

## Tidy Control Flow {.smaller}

::: {.columns .pv4}

::: {.column width="60%"}
- Part of what makes computers so useful is their ability to [change their behavior conditionally]{.b .green}
    - **If** [condition] is true... **then** do [behavior]
    
::: {.fragment .mt1}
- Programming this is called [control flow]{.b .blue}
    - Basic control flow is built into R (no packages)
    - [Tidy]{.b .green} control flow is easier **while wrangling**
:::

::: {.fragment .mt1}
- There are two main tidy control flow functions
    - For two outcomes: `if_else()`
    - For many outcomes: `case_when()`
:::

:::

::: {.column .tc .pv4 width="40%"}
{{< li usjxhgom trigger=loop delay=3000 colors=secondary:#2a76dd class=rc >}}
:::

:::

## Tidy Control Flow Live Coding

```{r}
#| echo: true
#| eval: false
#| error: true
#| code-line-numbers: false

library(tidyverse)

# if_else to assign to one of two values conditionally

## simple example
x <- 6
if_else(
  condition = x > 10,
  true = "large",
  false = "small"
)

## Compressed
if_else(x > 10, "large", "small")

## Vectorized
x <- c(6, 16, 3)
if_else(x > 10, "large", "small")

## Nested
if_else(
  condition = x > 10,
  true = "large",
  false = if_else(x > 5, "medium", "small")
)

# case_when to assign to one of 3+ values conditionally

## simple example
x <- c(6, 16, 3)
case_when(
  x > 10 ~ "large",
  x > 5 ~ "medium",
  x <= 5 ~ "small"
)

## We can also use `.default =` for when no conditions are met
case_when(
  x > 10 ~ "large",
  x > 5 ~ "medium",
  .default = "small"
)

## Pitfall: ordering the conditions poorly
case_when(
  x > 5 ~ "medium", 
  x > 10 ~ "large",
  .default = "small"
)

# Working within mutate: simple examples 

## Example data
sitetemps <- 
  tibble(
    site = c(2, 1, 2, 3, 3, 3, 1, 1, 2),
    temp = c(-2.1, 38.6, 4.6, -0.2, 4.5, 0.6, 37.6, 36.7, 102.3)
  ) |> 
  print()

## if_else within mutate
sitetemps |> 
  mutate(
    metric = if_else(
      condition = site == 1,
      true = "F",
      false = "C"
    )
  )

## case_when within mutate
sitetemps |> 
  mutate(
    nation = case_when(
      site == 1 ~ "US",
      site == 2 ~ "DE",
      site == 3 ~ "FR",
      .default = NA
    )
  )

# Working within mutate: complex examples

## Conditionally transforming values
sitetemps |> 
  mutate(
    metric = if_else(site == 1, "F", "C"),
    tempC = if_else(
      condition = metric == "F",
      true = (temp - 30) * (5 / 9), 
      false = temp
    ),
    state = case_when(
      tempC < 0 ~ "solid",
      tempC < 100 ~ "liquid",
      tempC > 100 ~ "gas"
    )
  )
```

# Tidy Selection

## Tidy Selection {.smaller}

::: {.columns .pv4}

::: {.column width="60%"}
-   `filter()` lets us retain rows using criteria
    -   A version for columns would be convenient!

::: {.fragment .mt1}
-   This is essentially what tidy selection does
    -   We can `select()` columns using criteria
    -   It's like an extension of the `:` sugar
:::

::: {.fragment .mt1}
-   We will use selection helpers and predicates
    -   `starts_with()`, `contains()`, etc.
    -   `everything()`, `last_col()`
    -   `where()` and `is.*()`

:::
:::

::: {.column .tc .pv4 width="40%"}
{{< li tvyxmjyo trigger=loop delay=3000 colors=secondary:#2a76dd class=rc >}}
:::

:::

## Tidy Selection Live Coding

```{r}
#| echo: true
#| eval: false
#| error: true
#| code-line-numbers: false

library(palmerpenguins)

penguins

# Starts with

penguins |> 
  select(starts_with("bill_"))

# PITFALL

penguins |> 
  starts_with("bill_") # error - don't forget select()
  

# Ends with

penguins |> 
  select(ends_with("_mm"))

# Contains

penguins |> 
  select(contains("_length_"))

# Where

penguins |> 
  select(where(is.factor))

penguins |> 
  select(where(is.numeric))

penguins |> 
  select(where(is.character))

# PITFALL

penguins |> 
  select(where(is.factor())) # error - don't include () after is.*

# Everything

penguins |> 
  select(everything())

# Last Column

penguins |> 
  select(last_col())

penguins |> 
  select(body_mass_g:last_col())

# Negating Criteria

penguins |> 
  select(!ends_with("_mm"))

# Combining Criteria

penguins |> 
  select(where(is.factor) | starts_with("body_"))

penguins |> 
  select(species, where(is.numeric))

penguins |> 
  select(ends_with("_mm") & !contains("_depth_"))
```

# Iterating Across

## Iterating Across {.smaller}

::: {.columns .pv4}

::: {.column width="60%"}
-   We can apply functions to many columns at once
    -   This is done using the `across()` function
    -   It goes inside `mutate()` and `summarize()`

::: {.fragment .mt1}
-   `across()` pairs very well with tidy selection
    -   e.g., do this to variables starting with "x"
    -   e.g., do this to all numeric variables
:::

::: {.fragment .mt1}
-   Advanced: "Anonymous functions"
    -   Call current col x, then round x to 2 digits
    -   e.g., `\(x) round(x, digits = 2)`
:::
:::

::: {.column .tc .pv4 width="40%"}
{{< li krmfspeu trigger=loop delay=3000 colors=secondary:#2a76dd class=rc >}}
:::

:::

## Across Live Coding

```{r}
#| echo: true
#| eval: false
#| error: true
#| code-line-numbers: false

starwars

# Applying the same mutation to multiple variables is a pain
starwars |> 
  mutate(
    hair_color = factor(hair_color),
    skin_color = factor(skin_color),
    eye_color = factor(eye_color),
  )

# dplyr 1.0.0 added across() to make this easier

starwars |> 
  mutate(
    across(
      .cols = c(hair_color, skin_color, eye_color), 
      .fns = factor
    )
  )

# We can also use tidyverse shortcuts like the colon (:)
starwars |> 
  mutate(
    across(
      .cols = c(hair_color:eye_color), 
      .fns = factor
    )
  )

# We can also compress it to take up less space
starwars |> 
  mutate(across(c(hair_color:eye_color), factor))

# We can also use a tidy selection helper when appropriate
starwars |> 
  mutate(
    across(
      .cols = ends_with("color"), 
      .fns = factor
    )
  )

starwars |>  
  mutate(across(.cols = where(is.character), .fns = factor))

# For more control, we use an anonymous function

starwars |> 
  mutate(
    height = height * 1000,
    mass = mass * 1000
  )

starwars |> 
  mutate(
    across(
      .cols = c(height, mass), 
      .fns = \(x) x * 1000
    )
  )

# Across with summarize

starwars |> 
  summarize(
    height = mean(height, na.rm = TRUE),
    mass = mean(mass, na.rm = TRUE),
    birth_year = mean(birth_year, na.rm = TRUE)
  )

starwars |> 
  summarize(
    across(
      .cols = c(height, mass, birth_year), 
      .fns = \(x) mean(x, na.rm = TRUE)
    )
  )
```
